---
title: "Importing data for number of people below/above poverty for language categories"
output:
  html_document:
    df_print: paged
---

```{r}
library(tidyverse)
library(tidycensus)
library(janitor)
```


Importing poverty data for each tx county for year 2010. I got this data from data.census.gov.

Link: https://data.census.gov/cedsci/table?q=poverty%20language&table=C16009&tid=ACSDT1Y2017.C16009&t=Poverty%3APoverty&hidePreview=true&g=0400000US48

```{r}
income_10 <- read_csv("data-income-poverty-counties/income_10.csv", skip = 1) %>% clean_names()

income_10

```

Let's look at Travis County. just for fun

```{r}
income_10 %>% 
  filter(geographic_area_name == "Travis County, Texas")
```

Okay.. looks like the margin of error for all counties is a little too big. Let's just try to do for Texas.


```{r}
income_tx_10 <- read_csv("data-income-poverty/tx_2010.csv", skip=1) %>% clean_names()

income_tx_10 %>% head()
```

cool. let's do it for all years. trying to use a function instead of importing one by one and binding later.

```{r}
data_path <- "data-income-poverty/"   # path to the data
files <- dir(data_path, pattern = "*.csv") # get file names

files

tx_poverty_language <- files %>%
  # read in all the files, appending the path before the filename
  map(~ read_csv(file.path(data_path, .), skip = 1)) %>% 
  reduce(rbind)

tx_poverty_language <- data_frame(filename = files) %>% # create a data frame
                                         # holding the file names
  mutate(file_contents = map(filename,          # read files into
           ~ read_csv(file.path(data_path, .), skip =1)) # a new data column
        )  

tx_poverty_language <- unnest(tx_poverty_language) %>% clean_names()


```

It worked! I got it from https://serialmentor.com/blog/2016/6/13/reading-and-combining-many-tidy-data-files-in-R

Cool. now let's clean the names.

```{r}
tx_poverty_language %>% names()
```

```{r}
names(tx_poverty_language) <- gsub("total_income_in_the_past_12_months_", "", names(tx_poverty_language))

tx_poverty_language <-tx_poverty_language %>% 
  mutate(
    filename = str_replace(filename, "tx_", ""),
    filename = str_replace(filename, ".csv", "")
  ) %>% 
  rename(year = filename)

tx_poverty_language

```

OK now let's export the data.

```{r}
saveRDS(tx_poverty_language, "data-out/tx_poverty_language.rds")

```
