---
title: "Exploring poverty by language spoken at home"
output:
  html_document:
    df_print: paged
---

```{r}
library(tidyverse)
library(tidycensus)
library(janitor)
```

```{r}
income_poverty <- readRDS("data-out/tx_poverty_language.rds")

income_poverty
```


#Questions

My questions for this dataset:
* What is the percentage of people above and below poverty for each language group? (visualize this.. pie chart? waffle chart?) probably only for 2017
* Is there a larger share of people below poverty within a certain language group?
* Is there a change overtime? (line chart)

Other questions:
* Is there a difference between income of those who speak only English vs speak other languages? (still need to find the dataset)

reminder: always take into account margin of error
always compare percentages, not number. certain language groups have more people. there are definitely more Spanish speakers than speakers of Asian languages.


I need to clean column names to make them easier to read

```{r}
  colnames(income_poverty) = gsub("margin_of_error", "moe", colnames(income_poverty))
  colnames(income_poverty) = gsub("speak_", "", colnames(income_poverty))
  colnames(income_poverty) = gsub("estimate_", "", colnames(income_poverty))
  
  income_poverty
  
```

Now I need to separate the data for above and below poverty level.

#At or Above Poverty Level

```{r}
above <- income_poverty %>% 
select(year, 
    contains("total"),
    contains("above")
  )

#exporting to use in next notebook
above %>% 
  saveRDS("data-out/above-1.rds")

```



Now I need to reshape my dataset so that the language is a separate column. Using gather()

```{r}
above <- above %>% 
  gather(key = "languages", "estimate", 6:15) %>% 
  arrange(year)

#need to get moe at a separate column. I think I can do this by filtering moe, separating it and then joining them? a little verbose but might be the only way..
moe_above <- above %>% 
  filter (str_detect(languages, "moe"))

moe_above <- moe_above %>% 
  rename(moe = estimate) %>% 
  mutate(languages = str_replace(languages, "moe_", "")) %>% 
  select(languages, moe, year)

above <- above %>% 
  filter (!str_detect(languages, "moe"))


#let's join them back
above_gathered <- left_join(above, moe_above, by = c("languages", "year"))

above_gathered


```

I think I need to clean the languages column 

```{r}
above_gathered <- above_gathered %>% 
  mutate(languages = str_replace(languages, "at_or_above_poverty_level_", "")) 

```

Exporting this data for use in next notebook

```{r}
above_gathered %>% 
  saveRDS("data-out/above_gathered.rds")
```

Now I want to do the same thing for below poverty

```{r}
below <- income_poverty %>% 
select(year, 
    contains("total"),
    contains("below")
  )

#exporting to use in next notebook
below %>% 
  saveRDS("data-out/below-1.rds")
  
```

```{r}
below <- below %>% 
  gather(key = "languages", "estimate", 6:15) %>% 
  arrange(year)

#need to get moe at a separate column. I think I can do this by filtering moe, separating it and then joining them? a little verbose but might be the only way..
moe_below <- below %>% 
  filter (str_detect(languages, "moe"))


moe_below <- moe_below %>% 
  rename(moe = estimate) %>% 
  mutate(languages = str_replace(languages, "moe_", "")) %>% 
  select(languages, moe, year)

below <- below %>% 
  filter (!str_detect(languages, "moe"))


#let's join them back
below_gathered <- left_join(below, moe_below, by = c("languages", "year"))

below_gathered <-below_gathered %>% 
  mutate(languages = str_replace(languages, "below_poverty_level_", "")) 


```

```{r}
below_gathered %>% 
  saveRDS("data-out/below_gathered.rds")
```

Now that I've separated the two so that I can use them in other notebooks, I actually need to join them back because I want to create percentage of the population below and at/above poverty level. So let's join the two nicely cleaned datasets again..

```{r}
above_below <- inner_join(above_gathered, below_gathered, by = c("total", "moe_total", "languages", "year"))

above_below <- above_below %>% 
  rename(estimate_above = estimate.x,
         moe_above = moe.x,
         estimate_below = estimate.y,
         moe_below = moe.y)


above_below
```


#Plotting percentage of population living below poverty level overtime

First, I need to create new columns for percentages of all speakers of that language

```{r}
above_below_pct <- above_below %>% 
  mutate(
    pct_below = estimate_below/(estimate_below+estimate_above)*100,
    pct_above = estimate_above/(estimate_below+estimate_above)*100
  )

above_below_pct
```
 
Plotting for below poverty.

```{r}
above_below_pct %>% 
  ggplot(aes(x = year, y = pct_below, group = languages)) +
  geom_line(aes(color = languages)) +
  labs(title = "Percentage of Texans with income below poverty level", subtitle = "based on languages spoken at home", caption = "Source: ACS 1-Year estimate", x = "Year", y = "% of total Population")
```


It seems that the % of the population living below poverty has declined in general for almost all groups. But here you can clearly see a gap between spanish speakers vs speakers of other language groups in terms of poverty.


There's probably no use in plotting for above poverty because it'll just be reversed.

```{r}
#exporting to use later
above_below_pct %>% saveRDS("data-out/above_below_pct.rds") 
```



#Are my questions for this dataset answered?
 
* What is the percentage of people above and below poverty for each language group? 
-sort of answered. But I'll create waffle charts to better illustrate this in the next notebook.
* Is there a larger share of people below poverty within a certain language group? 
-Yes. Spanish speakers.
* Is there a change overtime? 
-yes. poverty has generally declined.

 More questions:
 * Are Spanish speakers "more likely to be in poverty" or are people below poverty level more likely to be Spanish speakers?
 
 Who should I contact?
  - Nelson Flores, a professor at the University of Pennsylvania's Graduate School of Education - NPR article (https://www.npr.org/sections/ed/2017/09/15/545629043/research-shows-spanish-speakers-take-longer-to-learn-english-why)
  - Donald Hernandez
  https://cdlc.uncg.edu/wp-content/uploads/2012/12/Hernandez-etal-2009-School-age-children-in-immigrant-families.txt