---
title: "Waffle charts of language speakers in Texas & their poverty levels"
output:
  html_document:
    df_print: paged
---

```{r}
library(tidyverse)
library(janitor)
library(waffle)
```

I want to create a waffle chart for language speakers in Texas, and looking at the share of people below & at or above poverty level.

Reading data

```{r}
above <- readRDS("data-out/above-1.rds")
below <- readRDS("data-out/below-1.rds")

#renaming column names
names(above) <- names(above) %>% 
  str_replace_all("at_or_above_poverty_level_", "")

names(below) <- names(below) %>% 
  str_replace_all("below_poverty_level_", "")

below
```

Let's select for only the estimates and filter for only 2017

```{r}
above_2017 <- above %>% 
  filter(year == 2017) %>% 
  select(only_english, spanish, other_indo_european_languages, asian_and_pacific_island_languages, other_languages)

above_2017
```


```{r}
below_2017 <- below %>% 
  filter(year == 2017) %>% 
  select(only_english, spanish, other_indo_european_languages, asian_and_pacific_island_languages, other_languages)

total_2017 <- bind_rows(
  "below" = below_2017, 
  "above" = above_2017,
  .id = "poverty_status")

total_2017
```

Let's calculate sum of total language speakers (both below and above poverty)
```{r}
waffle_total_17 <- total_2017 %>% 
  select_if(is.numeric) %>%
   map_dbl(sum)
```

Now let's create a waffle chart!

```{r}
waffle(waffle_total_17/100000, rows=11, title = "Texas population by language spoken at home in 2017", xlab = "1 square is 100,000 people | source = ACS 1-Year estimates")
```

Spanish speaking population are a big part of the whole Texas population. Population speaking a language other than English at home makes up a third of the population.

Map spanish speakers

#Creating Waffle Chart for population above/below poverty, for all language groups.


first importing data from other notebook

```{r}
above_below <- readRDS("data-out/above_below_pct.rds")

above_below
```


```{r}
above_below_waffle <- above_below %>% 
  select (year, languages, pct_below, pct_above)

above_below_waffle_10 <- above_below_waffle %>% 
  filter(year == 2010) #filter for 2010, because I only want to create waffle charts for 2017

above_below_waffle_10
```

Now let's create waffle charts

```{r}
above_below_waffle_gathered_10 <- above_below_waffle_10 %>% 
  gather(key = "poverty_status", value="percentage", 3:4)

above_below_waffle_gathered_10
```

```{r}
english_10 <- above_below_waffle_10 %>% 
  mutate(year = NULL) %>% 
  filter(languages == "only_english") %>% 
  select(pct_above, pct_below)

spanish_10 <- above_below_waffle_10 %>% 
  mutate(year = NULL) %>% 
  filter(languages == "spanish") %>% 
  select(pct_above, pct_below)


european_10 <- above_below_waffle_10 %>% 
  mutate(year = NULL) %>% 
  filter(languages == "other_indo_european_languages") %>% 
  select(pct_above, pct_below)


asian_10 <- above_below_waffle_10 %>% 
  mutate(year = NULL) %>% 
  filter(languages == "asian_and_pacific_island_languages") %>% 
  select(pct_above, pct_below)


other_10 <- above_below_waffle_10 %>% 
  mutate(year = NULL) %>% 
  filter(languages == "other_languages") %>% 
  select(pct_above, pct_below)


```

creating multiple charts, combining with ggarange

```{r}
english_10

english <- waffle(english_10, rows=3, title = "English", colors = c("light blue", "maroon", "white"))

english

spanish <- waffle(spanish_10, rows=3, title = "Spanish", colors = c("light blue", "maroon", "white"))

asian <- waffle(asian_10, rows=3, title = "Asian and Pacific Island",  colors = c("light blue", "maroon", "white"))

european <- waffle(european_10, rows=3, title = "Indo-European", colors = c("light blue", "maroon", "white"))

other <- waffle(other_10, rows=3, title = "Other", colors = c("light blue", "maroon", "white"))

library(ggpubr)
waffle_2010 <- ggarrange(english, spanish, asian, european, nrow = 4, common.legend = TRUE, legend ="right")

waffle_2010 %>% annotate_figure(top = text_grob("Poverty Levels by Language Spoken at Home in 2010", color = "black", face = "bold", size = 20),
               bottom = text_grob("Data source: ACS 2010 1-Year Estimates", color = "black",
                                  hjust = 1, x = 1, face = "italic", size = 10)) 
  

```

Now I can see the difference in percentage of population living above and below poverty. Spanish speakers have a higher share of people living in poverty.

Now let's do it for 2017.

```{r}
above_below_waffle_17 <- above_below_waffle %>% 
  filter(year == 2017) #filter for 2010, because I only want to create waffle charts for 2017
above_below_waffle_gathered_17 <- above_below_waffle_10 %>% 
  gather(key = "poverty_status", value="percentage", 3:4)

english_17 <- above_below_waffle_17 %>% 
  mutate(year = NULL) %>% 
  filter(languages == "only_english") %>% 
  select(pct_above, pct_below)

spanish_17 <- above_below_waffle_17 %>% 
  mutate(year = NULL) %>% 
  filter(languages == "spanish") %>% 
  select(pct_above, pct_below)


european_17 <- above_below_waffle_17 %>% 
  mutate(year = NULL) %>% 
  filter(languages == "other_indo_european_languages") %>% 
  select(pct_above, pct_below)


asian_17 <- above_below_waffle_17 %>% 
  mutate(year = NULL) %>% 
  filter(languages == "asian_and_pacific_island_languages") %>% 
  select(pct_above, pct_below)


other_17 <- above_below_waffle_17 %>% 
  mutate(year = NULL) %>% 
  filter(languages == "other_languages") %>% 
  select(pct_above, pct_below)

```

Creating multiple charts for 2017

```{r}
english_2017 <- waffle(english_17, rows=3, title = "English", colors = c("light blue", "maroon", "white"))

spanish_2017 <- waffle(spanish_17, rows=3, title = "Spanish", colors = c("light blue", "maroon", "white"))

asian_2017 <- waffle(asian_17, rows=3, title = "Asian and Pacific Island languages",  colors = c("light blue", "maroon", "white"))

european_2017 <- waffle(european_17, rows=3, title = "Other Indo-European languages", colors = c("light blue", "maroon", "white"))

other_2017 <- waffle(other_17, rows=3, title = "Other", colors = c("light blue", "maroon", "white"))

library(ggpubr)
waffle_2017 <- ggarrange(english_2017, spanish_2017, asian_2017, european_2017, nrow = 4, common.legend = TRUE, legend ="right")

waffle_2017 %>% annotate_figure(top = text_grob("Poverty Levels by Language Spoken at Home in Texas", color = "black", face = "bold", size = 20),
               bottom = text_grob("Data source: ACS 2017 1-Year Estimates", color = "black",
                                  hjust = 1, x = 1, face = "italic", size = 10)) 
```

The gap is closer for 2017, compared with 2010.
