---
title: "Importing education levels"
output:
  html_document:
    df_print: paged
---

```{r}
library(tidyverse)
library(tidycensus)
```

I want to get census data for educational attainment based on language spoken at home. I looked at Factfinder and found: EDUCATIONAL ATTAINMENT AND EMPLOYMENT STATUS BY LANGUAGE SPOKEN AT HOME FOR THE POPULATION 25 YEARS AND OVER (code: B16010)


```{r}
v17 <- load_variables(2017, "acs5", cache = TRUE)

view(v17)
```




```{r}
language_by_education <- get_acs(geography = "state", variables = c("B16010_001", "B16010_002", "B16010_003", "B16010_004", "B16010_005", "B16010_006", "B16010_007", "B16010_008", "B16010_009", "B16010_010", "B16010_011", "B16010_012", "B16010_013", "B16010_014", "B16010_015", "B16010_016", "B16010_017", "B16010_018", "B16010_019", "B16010_020", "B16010_021", "B16010_022", "B16010_023", "B16010_024", "B16010_025", "B16010_026", "B16010_027", "B16010_028", "B16010_029", "B16010_030", "B16010_031", "B16010_032", "B16010_033", "B16010_034", "B16010_035", "B16010_036", "B16010_037", "B16010_038", "B16010_039", "B16010_040", "B16010_041", "B16010_042", "B16010_043", "B16010_044", "B16010_045", "B16010_046", "B16010_047", "B16010_048", "B16010_049", "B16010_050", "B16010_051", "B16010_052", "B16010_053"), year = 2017, state = "TX")

```


```{r}
language_by_education <- language_by_education %>% 
rename(
  name = variable,
  state = NAME
)

language_by_education <- inner_join(language_by_education, v17, by="name") 

language_by_education <- language_by_education %>% 
  mutate(concept = NULL)

language_by_education <- language_by_education %>% 
  mutate(
    label = str_replace(label, "Estimate!!Total!!", "")
  )

language_by_education <- language_by_education %>% 
  filter(str_detect(label, "Speak")) #filtering out the totals only

language_by_education

```

I need to separate the Label column

```{r}
language_by_education <- language_by_education %>% 
    separate(label, c("education", "employment", "language"), "!!")

language_by_education

```


great! exporting the data to use in another notebook. I want to visualize this with using waffle charts.

```{r}
language_by_education %>% saveRDS("data-out/language_by_education.rds")
```


