---
title: "Exploring education levels by language spoken at home"
output:
  html_document:
    df_print: paged
---

```{r}
library(tidyverse)
library(janitor)
library(waffle)
```


Importing data from other notebook

```{r}
language_by_education <- readRDS("data-out/language_by_education.rds")
```

```{r}
language_by_education
```


My goal in this notebook:
separating data by language groups and creating a waffle chart of education levels for each language group.

Filtering out for Spanish Only.


```{r}
spanish <- language_by_education %>% 
  filter(language == "Speak Spanish") %>% 
  select(language, education, employment, estimate, moe)

spanish 
```


up next: separate into education level only.

I'm also creating a new dataset that has percentage instead of count. will need this for creating waffle chart.
I'm also combining the numbers for both in the labor force or not in the labor force, because I'm not comparing that.

```{r}
spanish_by_education <- spanish %>% 
  group_by(education) %>% 
  summarise(sum=sum(estimate))


spanish_by_education <- spanish_by_education %>% 
  mutate(
    pct = sum / sum(sum) * 100
  )

spanish_pct <- spanish_by_education %>% 
  mutate(sum = NULL) %>% 
  spread(education, pct)
```

doing the same thing for English

```{r}
english <- language_by_education %>% 
  filter(language == "Speak only English")

english_by_education <- english %>% 
  group_by(education) %>% 
  summarise(sum=sum(estimate))

english_by_education <- english_by_education %>% 
  mutate(
    pct = sum / sum(sum) * 100
  )

english_pct <- english_by_education %>% 
  mutate(sum = NULL) %>% 
  spread(education, pct) %>% 
  select("Bachelor's degree or higher", "Some college or associate's degree", "High school graduate (includes equivalency)", "Less than high school graduate")

english_pct
```

doing the same thing for asian languages

```{r}
asian <- language_by_education %>% 
  filter(language == "Speak Asian and Pacific Island languages")

asian_by_education <- asian %>% 
  group_by(education) %>% 
  summarise(sum=sum(estimate))

asian_by_education <- asian_by_education %>% 
  mutate(
    pct = sum / sum(sum) * 100
  )

asian_pct <- asian_by_education %>% 
  mutate(sum = NULL) %>% 
  spread(education, pct) %>% 
  select("Bachelor's degree or higher", "Some college or associate's degree", "High school graduate (includes equivalency)", "Less than high school graduate")

asian_pct
```

doing the same thing for Indo-European languages


```{r}

indoeuropean <- language_by_education %>% 
  filter(language == "Speak other Indo-European languages")

indoeuropean_by_education <- indoeuropean %>% 
  group_by(education) %>% 
  summarise(sum=sum(estimate))

indoeuropean_by_education <- indoeuropean_by_education %>% 
  mutate(
    pct = sum / sum(sum) * 100
  )

indoeuropean_pct <- indoeuropean_by_education %>% 
  mutate(sum = NULL) %>% 
  spread(education, pct) %>% 
  select("Bachelor's degree or higher", "Some college or associate's degree", "High school graduate (includes equivalency)", "Less than high school graduate")
```

doing the same thing for other languages 

```{r}
other <- language_by_education %>% 
  filter(language == "Speak other languages")

other_by_education <- other %>% 
  group_by(education) %>% 
  summarise(sum=sum(estimate))

other_by_education <- other_by_education %>% 
  mutate(
    pct = sum / sum(sum) * 100
  )

other_pct <- other_by_education %>% 
  mutate(sum = NULL) %>% 
  spread(education, pct) %>% 
  select("Bachelor's degree or higher", "Some college or associate's degree", "High school graduate (includes equivalency)", "Less than high school graduate")
```

I will be creating separate charts and combining them with ggarrange.

```{r}
library(ggthemes)
library(ggplot2)

english_w <- waffle(english_pct, rows=3, colors = c("#ee4035", "#ffcc5c", "#7bc043", "#0392cf"), title = "English Only") + theme(legend.position = "none")

spanish_w <- waffle(spanish_pct, rows=3, colors = c("#ee4035", "#ffcc5c", "#7bc043", "#0392cf"), title ="Spanish") + theme(legend.position = "none")

spanish_w
indoeuropean_w <- waffle(indoeuropean_pct, rows=3, colors = c("#ee4035", "#ffcc5c", "#7bc043", "#0392cf"), title = "Other Indo-European languages") + theme(legend.position = "none")
asian_w <- waffle(asian_pct, rows=3, colors = c("#ee4035", "#ffcc5c", "#7bc043", "#0392cf"), title = "Asian & Pacific Island languages") + theme(legend.position = "none")

library(ggpubr)
combined_chart <- ggarrange(english_w, asian_w, indoeuropean_w, spanish_w, nrow = 4, common.legend = TRUE, legend ="right")


combined_chart

```

Adding a common title and source

```{r}
annotate_figure(combined_chart,
               top = text_grob("Educational Attainment by Language Spoken at Home", color = "black", face = "bold", size = 20),
               bottom = text_grob("Data source: ACS 2017 5-Year Estimates", color = "black",
                                  hjust = 1, x = 1, face = "italic", size = 10))
```


Conclusion:

Yes. A greater percentage of spanish speakers have low educational attainment, vs other language groups. How does this relate to poverty? I would need to ask my sources.
