---
title: "Importing Proficiency Data"
output:
  html_document:
    df_print: paged
---

```{r}
library(tidyverse)
library(janitor)
```


#Import

The last factor I want to compare between the four language speakers is proficiency. I downloaded this csv file from data.census.gov. This is the link to the data: https://data.census.gov/cedsci/table?q=languages%20spoken%20at%20home&table=S1601&tid=ACSST5Y2017.S1601&t=Language%20Spoken%20at%20Home&hidePreview=true&g=0400000US48


```{r}
proficiency_raw <- read.csv("data-raw/ACSST5Y2017.S1601_data_with_overlays_2019-05-08T214735.csv")

proficiency_raw 
```


This is a very messy dataset but I've noted from the key which columns I want. This dataset breaks the data down to age groups, which I don't need.

```{r}
proficiency_select <- proficiency_raw %>% 
  select("S1601_C03_004E", "S1601_C03_004M", "S1601_C03_008E", "S1601_C03_008M", "S1601_C03_012E", "S1601_C03_012M", "S1601_C03_016E", "S1601_C03_016M", "S1601_C05_004E", "S1601_C05_004M", "S1601_C05_008E", "S1601_C05_008M", "S1601_C05_012E", "S1601_C05_012M", "S1601_C05_016E", "S1601_C05_016M")

proficiency <- proficiency_select[-c(1),] #deleting the very ugly name row

proficiency

```

Now I have to rename these

for key: c03 means very well, c05 means less than very well
004 is Spanish
008 is other Indoeuropean languages
012 is Asian & Pacific Island languages
016 is other languages
E is estimate
M is margin of error

```{r}

proficiency <- proficiency %>% 
  rename(spanish_vwell = S1601_C03_004E,
         moe_spanish_vwell = S1601_C03_004M,
         indoeuropean_vwell = S1601_C03_008E,
         moe_indoeuropean_vwell = S1601_C03_008M,
         asian_vwell = S1601_C03_012E,
         moe_asian_vwell = S1601_C03_012M,
         other_vwell = S1601_C03_016E,
         moe_other_vwell = S1601_C03_016M,
         spanish_ltvwell = S1601_C05_004E,
         moe_spanish_ltvwell = S1601_C05_004M,
         indoeuropean_ltvwell = S1601_C05_008E,
         moe_indoeuropean_ltvwell = S1601_C05_008M,
         asian_ltvwell = S1601_C05_012E,
         moe_asian_ltvwell = S1601_C05_012M,
         other_ltvwell = S1601_C05_016E,
         moe_other_ltvwell = S1601_C05_016M)
```


Now I'll gather them 
```{r}
proficiency <- proficiency %>% 
  gather(key = "language", value = "estimate", 1:16)
```

Now I need to move the margin of error into a separate column. Totally long way of doing it...

```{r}
moe_table <- proficiency %>% 
  filter(str_detect(language, "moe"))

moe_table <- moe_table %>% 
  mutate(language = str_replace (language, "moe_", "")) %>% 
  rename(moe = estimate)

moe_table
```

Now doing the same thing for language table

```{r}
language_table <- proficiency %>% 
  filter(!str_detect(language, "moe"))

language_table 
```

Now joining

```{r}
language_proficiency <- inner_join(language_table, moe_table, by = "language")

language_proficiency
```

Let's make sure none of the MOE is more than 10% of the estimate

```{r}
language_proficiency <- language_proficiency %>% 
  mutate(moe = as.numeric(moe),
         estimate = as.numeric(estimate))#changing to numbers

language_proficiency %>% 
  mutate(pct_moe = moe/estimate * 100)
```

Looks like the moe is relatively small.

Good, now I need to separate estimate of very well and less than very well

```{r}
lep <- language_proficiency %>% 
  filter(str_detect(language, "ltvwell"))

lep <- lep %>% 
  mutate(language = str_replace(language, "_ltvwell", "")) %>% 
  rename(less_than_very_well = estimate,
         moe_less_than_very_well = moe)

lep

prof <- language_proficiency %>% 
  filter(!str_detect(language, "ltvwell"))

prof <- prof %>% 
  mutate(language = str_replace(language, "_vwell", "")) %>% 
  rename(very_well = estimate,
         moe_very_well = moe)

prof
```


Now I'll join the two.

```{r}
language_by_proficiency <- inner_join(lep, prof, by = "language")

language_by_proficiency
```

Now let's create new columns for percentages

```{r}
language_by_proficiency <- language_by_proficiency %>% 
  mutate(pct_very_well = very_well / (very_well + less_than_very_well) * 100,
         pct_less_than_very_well = less_than_very_well / (very_well + less_than_very_well) * 100)

language_by_proficiency
```

Exporting the clean data for use in another notebook

```{r}
language_by_proficiency %>% saveRDS("data-out/02_language_proficiency.rds")

```

