---
title: "Visualizing proficiency by languages spoken at home"
output:
  html_document:
    df_print: paged
---

```{r}
library(tidyverse)
library(ggplot2)
library(waffle)
library(janitor)
```

Importing from other notebook

```{r}
language_by_proficiency <- read_rds("data-out/02_language_proficiency.rds")

language_by_proficiency
```

Now I'm going to create waffle charts for each of the language group.

```{r}
language_proficiency_pct <- language_by_proficiency %>% 
  select(language, pct_very_well, pct_less_than_very_well)

language_proficiency_pct %>% 
  arrange(desc(pct_very_well))
```

Here you can see that the percentage speaking very well for Spanish speakers is slightly higher than for Asian languages.

Before creating waffle charts, I need to separate the data into the different language categories and delete the language column.

```{r}
spanish_p <- language_proficiency_pct %>% 
  filter(language == "spanish") %>% 
  mutate(language = NULL)

indoeuropean_p <- language_proficiency_pct %>% 
  filter(language == "indoeuropean") %>% 
  mutate(language = NULL)

asian_p <- language_proficiency_pct %>% 
  filter(language == "asian") %>% 
  mutate(language = NULL)

other_p <- language_proficiency_pct %>% 
  filter(language == "other") %>% 
  mutate(language = NULL)
```

Now charting

```{r}
spanish_pw <- waffle(spanish_p, rows=3, colors = c("#3caea3", "#ff5733"), title = "Spanish") + theme(legend.position = "none")

indoeuropean_pw <- waffle(indoeuropean_p, rows=3, colors = c("#3caea3", "#ff5733"), title = "Other Indo-European Languages") + theme(legend.position = "none")

asian_pw <- waffle(asian_p, rows=3, colors = c("#3caea3", "#ff5733"), title = "Asian & Pacific Island Languages") + theme(legend.position = "none")

other_pw <- waffle(other_p, rows=3, colors = c("#3caea3", "#ff5733"), title = "Other Languages") + theme(legend.position = "none")

library(ggpubr)
proficiency_chart <- ggarrange(spanish_pw, asian_pw, indoeuropean_pw, other_pw, nrow = 5, common.legend = TRUE, legend ="right")

annotate_figure(proficiency_chart,
               top = text_grob("Languages Spoken at Home by English Proficiency", color = "black", face = "bold", size = 20),
               bottom = text_grob("Data source: ACS 2017 5-Year Estimates", color = "black",
                                  hjust = 1, x = 1, face = "italic", size = 10))

```

Conclusion: At-home Spanish speakers have a slightly higher percentage of people speaking English very well compared to Asian & Pacific Island Languages speakers.

People speaking other Indo-European languages at home have a really high percentage of people speaking English very well.